---
title: "Problem Set 5"
author: "Elizabeth Bailey"
date: "9/29/2021"
output:
  word_document: default
  pdf_document: default
  html_document: default
---

data files converted to CSV files with these commands

echo 'YEAR, MONTH, DTEMPC' >! Hadcrut_GlobalAverage.csv
sed 's[/[ [' HadCRUT.4.6.0.0.monthly_ns_avg.txt | awk '{print $1 ", " $2 ", " $3}' >> Hadcrut_GlobalAverage.csv

echo 'YEAR, MONTH, DTEMPC' >! Hadcrut_NHAverage.csv
sed 's[/[ [' HadCRUT.4.6.0.0.monthly_nh.txt | awk '{print $1 ", " $2 ", " $3}' >> Hadcrut_NHAverage.csv

echo 'YEAR, MONTH, DTEMPC' >! Hadcrut_SHAverage.csv
sed 's[/[ [' HadCRUT.4.6.0.0.monthly_sh.txt | awk '{print $1 ", " $2 ", " $3}' >> Hadcrut_SHAverage.csv


Question 1a -- t-test on the first and last decades of the 20th century

```{r}

had_ns_avg <- read.csv("Hadcrut_GlobalAverage.csv")
temps <- had_ns_avg$DTEMPC
years <- had_ns_avg$YEAR
months <- had_ns_avg$MONTH

time <- years + (months-0.5)/12.0

had_sh_avg <- read.csv("Hadcrut_SHAverage.csv")
had_nh_avg <- read.csv("Hadcrut_NHAverage.csv")
temps_sh <- had_sh_avg$DTEMPC
temps_nh <- had_nh_avg$DTEMPC

N <- length(time)
dtime <- seq(1900,1999,by=10)
dN <- length(dtime)

decavg <- matrix(0,dN,1)
decvar <- matrix(0,dN,1)

tt <- time>= 1900 & time<1910
tem0 <- temps[tt == TRUE]
N0 <- length(tem0)
avg0 <- mean(tem0)
var0 <- var(tem0)

tt <- time>= 1990 & time<2000
tem1 <- temps[tt == TRUE]
N1 <- length(tem1)
avg1 <- mean(tem1)
var1 <- var(tem1)

t0 <- (sqrt(N0-1)*(avg0))/sqrt(var0)
t0
t1 <- (sqrt(N1-1)*(avg1))/sqrt(var1)
t1

print(paste("mean of 1900s is ",avg0," degrees C, with standard deviation ",sqrt(var0)))
print(paste("mean of 1990s is ",avg1," degrees C, with standard deviation ",sqrt(var1)))

```
Question 1b -- two-way T-tests of decadal sample means
```{r}

t <- (sqrt(N0-1)*(avg0-avg1))/sqrt(var0 + var1)
t

testt <- t.test(tem0, tem1)
print(testt$statistic)

```
Question 1c -- t.tests of sample means of successive decades

The temperatures increase for 10 successive decade pairs and decrease for 6 successive decade pairs. I think global warming began some time around the turn of the 20th century. 
```{r}

decade1_t <- 1:16
decade1_logp <- 1:16
decade2_t <- 1:16
decade2_logp <- 1:16
dtime <- seq(1860,2010,by=10)
dN <- length(decade1_t)

grt <- matrix(0,dN,1)
lss <- matrix(0,dN,1)
calct <- matrix(0,dN,1)
funt <- matrix(0,dN,1)
ptest <- matrix(0,dN,1)

k <- 0
for (i in seq(1850,2000,by=10)) {
  k <- k + 1
  t1 <- time>=i & time<(i+10)
  tem1 <- temps[t1 == TRUE]
  t2 <- time>=(i+10) & time<(i+20)
  tem2 <- temps[t2 == TRUE]
  N0 <- length(tem2)
  avg1 <- mean(tem1)
  avg2 <- mean(tem2)
  var1 <- var(tem1)
  var2 <- var(tem2)
  t <- (sqrt(N0-1)*(avg1-avg2))/sqrt(var1 + var2)
  calct[k] <- t
  testt <- t.test(tem1, tem2, alternative = "two.sided", conf.level = 0.95)
  funt[k] <- testt$statistic
  ptest[k] <- testt$p.value 
  great <- t.test(tem1, tem2, alternative = "greater", conf.level = 0.99)
  grt[k] <- great$p.value
  less <- t.test(tem1, tem2, alternative = "less", conf.level = 0.99)
  lss[k] <- less$p.value
}

g <- grt[grt==TRUE]
decades <- seq(1855,2014,by=10)

plot(decades, ptest, log = 'y')
title("Decadal comparisons: Global Average Temperature Anomalies")
plot(decades, grt)
title("Decadal comparisons: Global Average Temperature Anomalies (Greater)")
plot(decades, lss)
title("Decadal comparisons: Global Average Temperature Anomalies(Less)")

```

Question 1d -- t.tests of sample means of successive decades: Northern Hemisphere

The temperatures increase for 8 successive decade pairs and decrease for 7 successive decade pairs. I think global warming began some time around the 1910s. 
```{r}

# par(mfrow=c(2,2))

decade1_t <- 1:16
decade1_logp <- 1:16
decade2_t <- 1:16
decade2_logp <- 1:16
dtime <- seq(1860,2010,by=10)
dN <- length(decade1_t)

grt <- matrix(0,dN,1)
lss <- matrix(0,dN,1)
calct <- matrix(0,dN,1)
funt <- matrix(0,dN,1)
ptest <- matrix(0,dN,1)

k <- 0
for (i in seq(1850,2000,by=10)) {
  k <- k + 1
  t1 <- time>=i & time<(i+10)
  tem1 <- temps_nh[t1 == TRUE]
  t2 <- time>=(i+10) & time<(i+20)
  tem2 <- temps_nh[t2 == TRUE]
  N0 <- length(tem2)
  avg1 <- mean(tem1)
  avg2 <- mean(tem2)
  var1 <- var(tem1)
  var2 <- var(tem2)
  t <- (sqrt(N0-1)*(avg1-avg2))/sqrt(var1 + var2)
  calct[k] <- t
  testt <- t.test(tem1, tem2, alternative = "two.sided", conf.level = 0.95)
  funt[k] <- testt$statistic
  ptest[k] <- testt$p.value 
  great <- t.test(tem1, tem2, alternative = "greater", conf.level = 0.99)
  grt[k] <- great$p.value
  less <- t.test(tem1, tem2, alternative = "less", conf.level = 0.99)
  lss[k] <- less$p.value
}

g <- grt[grt==TRUE]
decades <- seq(1855,2014,by=10)

plot(decades, ptest, log = 'y')
title("Decadal comparisons: NH Average Temperature Anomalies")
plot(decades, grt)
title("Decadal comparisons: NH Average Temperature Anomalies (Greater)")
plot(decades, lss)
title("Decadal comparisons: NH Average Temperature Anomalies(Less)")

```

Question 1e -- t.tests of sample means of successive decades: Southern Hemisphere

The temperatures increase for 12 successive decade pairs and decrease for 4 successive decade pairs. I think global warming began some time around the turn of the 20th century. 
```{r}

decade1_t <- 1:16
decade1_logp <- 1:16
decade2_t <- 1:16
decade2_logp <- 1:16
dtime <- seq(1860,2010,by=10)
dN <- length(decade1_t)

grt <- matrix(0,dN,1)
lss <- matrix(0,dN,1)
calct <- matrix(0,dN,1)
funt <- matrix(0,dN,1)
ptest <- matrix(0,dN,1)

k <- 0
for (i in seq(1850,2000,by=10)) {
  k <- k + 1
  t1 <- time>=i & time<(i+10)
  tem1 <- temps_sh[t1 == TRUE]
  t2 <- time>=(i+10) & time<(i+20)
  tem2 <- temps_sh[t2 == TRUE]
  N0 <- length(tem2)
  avg1 <- mean(tem1)
  avg2 <- mean(tem2)
  var1 <- var(tem1)
  var2 <- var(tem2)
  t <- (sqrt(N0-1)*(avg1-avg2))/sqrt(var1 + var2)
  calct[k] <- t
  testt <- t.test(tem1, tem2, alternative = "two.sided", conf.level = 0.95)
  funt[k] <- testt$statistic
  ptest[k] <- testt$p.value 
  great <- t.test(tem1, tem2, alternative = "greater", conf.level = 0.99)
  grt[k] <- great$p.value
  less <- t.test(tem1, tem2, alternative = "less", conf.level = 0.99)
  lss[k] <- less$p.value
}

g <- grt[grt==TRUE]
decades <- seq(1855,2014,by=10)

plot(decades, ptest, log = 'y')
title("Decadal comparisons: SH Average Temperature Anomalies")
plot(decades, grt)
title("Decadal comparisons: SH Average Temperature Anomalies (Greater)")
plot(decades, lss)
title("Decadal comparisons: SH Average Temperature Anomalies(Less)")

```

Question 2a -- Plot some F variance ratio distributions

```{r}

fvarrat <- seq(.001,10.0,by=0.001)
f1 <- df(fvarrat, df1 = 2, df2 = 8, ncp = 0)
f2 <- df(fvarrat, df1 = 10, df2 = 20, ncp = 0)
f3 <- df(fvarrat, df1 = 17, df2 = 2000, ncp = 0)

plot(f3, type = 'l', col = 'blue')
lines(f2, type = 'l', col = 'green')
lines(f1, type = 'l', col = 'red')
title(" 2 & 8 (red), 10 & 20 (green) and 17 & 2000 (blue) dof",ylab="F-distribution PDF")

f1 <- df(fvarrat, df1 = 2, df2 = 8, ncp = 0, log = TRUE)
f2 <- df(fvarrat, df1 = 10, df2 = 20, ncp = 0, log = TRUE)
f3 <- df(fvarrat, df1 = 17, df2 = 2000, ncp = 0, log = TRUE)

plot(f3, type = 'l', col = 'blue')
lines(f2, type = 'l', col = 'green')
lines(f1, type = 'l', col = 'red')
title(" 2 & 8 (red), 10 & 20 (green) and 17 & 2000 (blue) dof",ylab="LOG F-distribution PDF")

```

Question 2b -- ANOVA with global temperature anomalies

```{r}

N <- 17
decade_means <- matrix(0,N,1)
decade_vars <- matrix(0,N,1)
totpts <- 17*120

k=0
for (i in seq(1850,2019,by=10)) {
  k = k + 1
  tt <- time>=i & time<(i+10)
  tem <- temps[tt == TRUE]
  men <- mean(tem)
  decade_means[k] <- men
  ver <- var(tem)
  decade_vars[k] <- ver
}
grandmen <- mean(temps)

sum1 <- 0 
for (i in 1:N) {
  squr <- (decade_means[i]-grandmen)^2
  sum1 <- sum1 + squr
}
Yb <- (1/(N-1))*sum1

M <- 120
dec <- 1850
sum2 <- 0
mfrac <- 1/(M-1)
for (j in 1:N) {
  sum3 <- 0
  for (i in 1:M){
    tt <- time>=dec & time<(dec+10)
    tem <- temps[tt == TRUE]
    sqar <- (tem[i]-decade_means[j])^2
    sum3 <- sum3 + sqar
  }
  mul <- mfrac*sum3
  sum2 <- sum2 + mul
  dec <- dec+10
}
Yw <- (1/N)*sum2

# M <- 120
# dec <- 1850
# sum2 <- 0
# mfrac <- 1/(M-1)
# for (j in 1:N) {
#   tt <- time>=dec & time<(dec+10)
#   tem <- temps[tt == TRUE]
#   sqar <- var(tem)
#   mul <- mfrac*sqar
#   sum2 <- sum2 + mul
#   dec <- dec+10
# }
# Yw1 <- (1/N)*sum2

FF <- Yb/Yw
FF

f <- round(FF,4)

d1 <- N-1
d2 <- N*(M-1)
fcdf <- pf(fvarrat, df1 = d1, df2 = d2)
plot(fvarrat, fcdf, type = 'l')

ff <- fvarrat>(f-.001) &fvarrat<(f+.001)

anova_prob <- fcdf[ff==T]
pval <- 1-anova_prob
print(paste("ANOVA probability of nonrandomness is ",anova_prob,"  p-value: ",pval))

```

Question 2c -- slice the temp anomalies by month

```{r}

n <- length(temps)

N <- 12
M <- 120
monthly_means <- matrix(0,N,1)
monthly_vars <- matrix(0,N,1)
totpts <- 17*120

for (i in 1:N) {
  mon <- seq(i, n, by=12)
  tem <- temps[mon]
  men <- mean(tem[1:M])
  monthly_means[i] <- men
  ver <- var(tem[1:M])
  monthly_vars[i] <- ver
}

tttp <- years >= 1850 & years<=1970

grandmen <- mean(temps[tttp==TRUE])

sum1 <- 0 
for (i in 1:N) {
  squr <- (monthly_means[i]-grandmen)^2
  sum1 <- sum1 + squr
}
Yb <- (1/(N-1))*sum1

sum2 <- 0
mfrac <- 1/(M-1)
for (j in 1:N) {
  sum3 <- 0
  mon <- seq(j, n, by=12)
  tem <- temps[mon]
  for (i in 1:M){
    sqar <- (tem[i]-monthly_means[j])^2
    sum3 <- sum3 + sqar
  }
  mul <- mfrac*sum3
  sum2 <- sum2 + mul
}
Yw <- (1/N)*sum2

FF <- Yb/Yw
FF

d1 <- N-1
d2 <- N*(M-1) 
fcdf <- pf(fvarrat, df1 = d1, df2 = d2)
plot(fvarrat, fcdf, type = 'l')

print(paste("stdev of within is ",Yw,"  stdev of among is ",Yb))
print(paste("F variance ratio is ",FF))

f <- round(FF,3)
ff <- fvarrat>(f-.001) & fvarrat<(f+.001)

testt <- t.test(temps)

anova_prob <- fcdf[ff==T]
pval <- 1-anova_prob

print(paste("ANOVA probability of nonrandomness is ",anova_prob,"  p-value: ",pval))

```

Question 3a -- the correlation of hemispheric temperature anomalies

```{r}

N <- 120
nn <- 17
r <- matrix(0,nn,1)

k <- 0
for (i in seq(1850,2019,by=10)) {
  k = k + 1
  tt <- time>=i & time<(i+10)
  tem_sh <- temps_sh[tt == TRUE]
  tem_nh <- temps_nh[tt == TRUE]
  men_sh <- mean(tem_sh)
  men_nh <- mean(tem_nh)
  sum1 <- 0 
  sum2 <- 0
  sum3 <- 0
  for (j in 1:N){
    diff_sh <- tem_sh[j] - men_sh
    diff_nh <- tem_nh[j] - men_nh
    num <- diff_sh*diff_nh
    sum1 <- sum1 + num
    sqar_sh <- diff_sh^2
    sum2 <- sum2 + sqar_sh
    sqar_nh <- diff_nh^2
    sum3 <- sum3 + sqar_nh
  }
  corr <- sum1/sqrt(sum2*sum3)
  r[k] <- corr
}

decades <- seq(1855,2019,by=10)
decades

plot(decades, r, type='l')

```

Question 3b -- the correlation of hemispheric temperature anomalies

```{r}
x <- (N-2)*(r^2)
d1 <- 1
d2 <- N-2

fcdf <- pf(x, df1 = d1, df2 = d2)
plot(x, fcdf)
# plot(dtime,pvals)

```
Question 3c

```{r}

prb <- 1-fcdf
plot(prb, log = 'y', type = 'l')

```

Question 3d -- compute the correlation between the sequences of decadal means for the Northern and Southern Hemispheres
  
```{r}

N <- 120
nn <- 17

men_sh_tot <- mean(temps_sh)
men_nh_tot <- mean(temps_nh)

sum1 <- 0 
sum2 <- 0
sum3 <- 0
for (i in seq(1850,2019,by=10)) {
  tt <- time>=i & time<(i+10)
  tem_sh <- temps_sh[tt == TRUE]
  tem_nh <- temps_nh[tt == TRUE]
  men_sh <- mean(tem_sh)
  men_nh <- mean(tem_nh)
  diff_sh <- men_sh - men_sh_tot
  diff_nh <- men_nh - men_nh_tot
  num <- diff_sh*diff_nh
  sum1 <- sum1 + num
  sqar_sh <- diff_sh^2
  sum2 <- sum2 + sqar_sh
  sqar_nh <- diff_nh^2
  sum3 <- sum3 + sqar_nh
}

corr <- sum1/sqrt(sum2*sum3)
r <- corr
r

ff <- (nn-1)*r
ff

d1 <- 1
d2 <- nn-2

fcdf <- pf(ff, df1 = d1, df2 = d2)
prob <- 1 - fcdf
prob

```