---
title: "Problem Set 4"
author: "Elizabeth Bailey"
date: "9/23/2021"
output:
  word_document: default
  pdf_document: default
  html_document: default
---

Quesiton 1


```{r}

had_ns_avg <- read.csv("Hadcrut_GlobalAverage.csv")
names(had_ns_avg)
temps <- had_ns_avg$DTEMPC
length(temps)
years <- had_ns_avg$YEAR
months <- had_ns_avg$MONTH

time <- years + (months-0.5)/12.0
time[1:24]
 
```
Question 1a

```{r}
N <- length(temps)
runavg <- matrix(0,N-5,1)

for (k in 5:N) {
  sum <- 0
  for (j in -5:5) {
    sum <- sum + temps[k+j]
  }
  avg <- (1/11)*sum
  runavg[k-5] <- avg
}
head(runavg)

rnt <- time[6:N]

plot(time, temps)
lines(rnt, runavg, col='lightgreen')
title("Hadley Center Global Average Temperature Anomalies",xlab="year",ylab="°C anomaly referenced to 1961-1990 average")


```

Question 1b -- sample mean and sample variance of entire time series

```{r}

men <- sum(temps)/N
men
mean(temps)

dif <- (temps-men)
difsq <- dif^2
vary <- sum(difsq)*(1/(N-1))
vary

var(temps)

x <- seq(min(temps), max(temps), by=.01)
pfx <- dnorm(x, mean = men, sd = sqrt(vary))
gss <- (pfx*N)/19

hist(temps, breaks = 50)
lines(x,gss)


```

Question 1c  -- QQ plot

```{r}

int <- (max(temps) - min(temps))/N


x <- seq(min(temps)+int, max(temps), by=int)
length(x)

empcdf <- ecdf(temps)
ecf <- empcdf(x)
plot(x, ecf, type='l')
title('Empirical and Theoretical CDFs')

tcdf <- pnorm(x, mean = men, sd = sqrt(vary))
lines(x, tcdf, type='l')

u <- seq(0, 1, by=.001)
v <- seq(0, 1, by=.001)
plot(tcdf, ecf, type='l',xlab="Empirical CDF",ylab="Gaussian CDF")
lines(u,v, type='l', col='gray')
title("QQ Plot of Global-Average Temp Anomalies")

ldevi <- 0
for (i in 1:N) {
  dff <- abs(ecf[i]-tcdf[i])
  if (dff > ldevi) {
    ldevi <- dff
  }
}

ldevi

```
Question 1d  -- Beta distribution and the median

```{r}
N <- length(temps)
length(ecf)
med <- median(temps)
med
# medtemp <- which(temps==med)
# medtemp
# temps[medtemp==TRUE]
e <- pnorm(med, mean = men, sd = sqrt(vary))
e
# e5 <- 1-ecf==0.5
# i <- temps[e5==TRUE]
# i
ro <- e - 0.5
fact <- (1-(4*(ro)^2))^(N/2)
fact
# print(paste("the CDF deviation of the median (\rho) is ",rho))

# print(paste("the Beta PDF at the sample median is ",pdf_factor," smaller than at the theoretical median"))

```

Question 1e -- sample means of decades

Yes, there is an apparent positive trend in these plots, indicating global wamring.
```{r}

N <- length(time)
decade_means <- 1:17
dN <- length(decade_means)
dtime <- seq(1855,2015,by=10)


decavg <- matrix(0,dN,1)
decvar <- matrix(0,dN,1)
k=0
for (i in seq(1850,2019,by=10)) {
  k = k + 1
  tt <- time>=i & time<(i+10)
  tem <- temps[tt == TRUE]
  men <- mean(tem)
  decavg[k] <- men
  ver <- var(tem)
  decvar[k] <- ver
}
decavg
st1 <- decavg - sqrt(decvar)
st2 <- decavg + sqrt(decvar)

plot(dtime, decavg, type='l')
arrows(dtime, st1, dtime, st2, length = 0)
title("Hadley Center Global Average Temperature Anomalies")

plot(dtime, decvar, type='l')

```
Problem 1f -- compare first and last decades of the 20th century

Global warming is apparent when looking at the shift of mean in these plots.
```{r}

N <- length(time)
dtime <- seq(1900,1999,by=10)
dN <- length(dtime)

decavg <- matrix(0,dN,1)
decvar <- matrix(0,dN,1)

tt <- time>= 1900 & time<1910
tem0 <- temps[tt == TRUE]
men <- mean(tem0)
avg0 <- men
ver <- var(tem0)
var0 <- ver

tt <- time>= 1990 & time<2000
tem1 <- temps[tt == TRUE]
men <- mean(tem1)
avg1 <- men
ver <- var(tem1)
var1 <- ver


min <- min(temps)
max <- max(temps)

x <- seq(min, max, by = 0.01)

pf <- dnorm(x, mean = avg0, sd = sqrt(var0))
cf <- pnorm(x, mean = avg0, sd = sqrt(var0))
ef <- ecdf(tem0)
emp <- ef(x)

pf2 <- dnorm(x, mean = avg1, sd = sqrt(var1))
cf2 <- pnorm(x, mean = avg1, sd = sqrt(var1))
ef2 <- ecdf(tem1)
emp2 <- ef2(x)

plot(x, pf, type='l')
title("1900 to 1909")
plot(x, pf2, type='l')
title("1990 to 1999")
plot(emp, cf, type = 'l')
plot(emp2, cf2, type = 'l')

plot(x, cf, type = 'l')
lines(x, cf2, type = 'l')
lines(x, emp, type = 'l')
lines(x, emp2, type = 'l')
# title("first and last decades of 20th century Global-Average Temp Anomalies",xlab="Degrees Celsius",ylab="Empirical and Gaussian CDFs")



```
Problem 2 -- Earthquake distributions

Question 2c

```{r}

gevents <- read.csv("GlobalEarthquakes6.0.csv")
names(gevents)

year <- gevents$YEAR
mags <- gevents$MAG

nevents <- length(mags)
dstep <- 1.0/nevents
steps <- seq(1.0-dstep/2.0,dstep/2.0,length.out = nevents)
# plot(mags,steps,type="l",log="y")
minn <- min(mags)
maxx <- max(mags)
int <- (maxx-minn)/nevents
int
x <- seq(minn, maxx, by=int)

ef <- ecdf(mags)
emp <- ef(x)

plot(x, emp, type='l')

```
Question 2d

```{r}
y <- 1 - emp
plot(x, y, type = 'l', log="y")
```

Question 2e -- the frequency of great earthquakes versus Poisson distribution

Yes, great earthquakes follow the poisson distribution. 

```{r}
yrs <- 1980:2020
yn <- length(yrs)
sum <- 0
ma <- matrix(0, yn, 1)
ys <- matrix(0, yn, 1)

j <- 0
for (i in yrs) {
  j <- j + 1
  yr <- year==i
  m <- mags[yr==TRUE]
  m8 <- m>=8.0
  m8true <- m[m8==TRUE]
  ml <- length(m8true)
  ma[j] <- ml
  ys[j] <- i
}

ma
sortm <- sort(ma)
n <- length(sortm)
lamb <- mean(ma)
int <- (max(ma)-min(ma))/n
k <- seq(min(ma), max(ma), by=1)

ppdf <- dpois(k, lambda = lamb)
emppdf <- dpois(sortm, lambda = lamb)

pcdf <- ppois(k, lambda = lamb)

plot(k, ppdf, type = 'l', col ='red')
lines(sortm, emppdf, type = 'l')

plot(k, pcdf, type = 'l')

```

Question 2f -- the frequency of very large earthquakes versus Poisson distribution  

Yes, very large earthquakes follow the poisson distribution.

```{r}

yrs <- 1980:2020
yn <- length(yrs)
sum <- 0
ma <- matrix(0, yn, 1)
ys <- matrix(0, yn, 1)

j <- 0
for (i in yrs) {
  j <- j + 1
  yr <- year==i
  m <- mags[yr==TRUE]
  m8 <- m>=7.5
  m8true <- m[m8==TRUE]
  ml <- length(m8true)
  ma[j] <- ml
  ys[j] <- i
}
ma

sortm <- sort(ma)
n <- length(sortm)
lamb <- mean(ma)
int <- (max(ma)-min(ma))/n
k <- seq(min(ma), max(ma), by=1)

ppdf <- dpois(k, lambda = lamb)
emppdf <- dpois(sortm, lambda = lamb)

pcdf <- ppois(k, lambda = lamb)

plot(k, ppdf, type = 'l', col ='red')
plot(sortm, emppdf, type = 'l')

plot(k, pcdf, type = 'l')

```

