---
title: "Problem Set 7"
author: "Elizabeth Bailey"
date: "10/13/2021"
output:
  word_document: default
  pdf_document: default
  html_document: default
---

data files converted to CSV files with these commands

echo 'YEAR, MONTH, TIME, CO2 ' >! co2_maunaloa.csv
awk 'NR>53 {print $1 ", " $2 ", " $3 ", " $4}' co2_mm_mlo.txt >> co2_maunaloa.csv

Problem 1

```{r}

mlo_co2 <- read.csv("co2_maunaloa.csv")
names(mlo_co2)
temps <- mlo_co2$DTEMPC
length(temps)
year <- mlo_co2$YEAR
months <- mlo_co2$MONTH
time <- mlo_co2$TIME
co2 <- mlo_co2$CO2


plot(time,co2,col='black',type="l",xlab="",ylab="")
title("Mauna Loa Carbon Dioxide Measurements",xlab="year",ylab="°Parts Per Million")

```

Problem 1a --  fit to Mauna Loa CO2 in decades

They increase over time. At 1980, it transition to a more constant amplitude. 
```{r}

# pre-define arrays for regression parameters of seven data segments
ann1 <- 1:7
ann1err <- 1:7
ann2 <- 1:7
ann2err <- 1:7
yr <- 1:7

pi2 <- 2*pi
pi4 <- 4*pi
anns <- sin(pi2*time)
annc <- cos(pi2*time)
ann2s <- sin(pi4*time)
ann2c <- cos(pi4*time)
# initialize segment counter
k=0
# the data segments are defined by their start-points
for (j in c(1,109,217,321,429,537,637)){
  k=k+1
  timed <- time[j:(j+119)]
  time2d <- timed^2
  co2d <- co2[j:(j+119)]
  annsd <- anns[j:(j+119)]
  anncd <- annc[j:(j+119)]
  ann2sd <- ann2s[j:(j+119)]
  ann2cd <- ann2c[j:(j+119)]
  lmlmlm <- lm(co2d ~ timed + time2d + annsd + anncd + ann2sd + ann2cd)
  summary(lmlmlm)
  anns1 <- coef(summary(lmlmlm))["annsd","Estimate"]
  annc1 <- coef(summary(lmlmlm))["anncd","Estimate"]
  ann1[k] <- sqrt(anns1^2+annc1^2)
  anns1 <- coef(summary(lmlmlm))["annsd","Std. Error"]
  annc1 <- coef(summary(lmlmlm))["anncd","Std. Error"]
  ann1err[k] <- sqrt(anns1^2+annc1^2)
  anns2 <- coef(summary(lmlmlm))["ann2sd","Estimate"]
  annc2 <- coef(summary(lmlmlm))["ann2cd","Estimate"]
  ann2[k] <- sqrt(anns2^2+annc2^2) 
  yr[k] <- year[j+60]
  anns2 <- coef(summary(lmlmlm))["ann2sd","Std. Error"]
  annc2 <- coef(summary(lmlmlm))["ann2cd","Std. Error"]
  ann2err[k] <- sqrt(anns2^2+annc2^2)
}


plot(yr,ann1,col='black',xlab="Year",ylab="Annual Cycle Amplitude (ppm CO2)",ylim=range(ann1+ann1err,ann1-ann1err))
arrows(x0=yr,y0=ann1+ann1err,x1=yr,y1=ann1-ann1err,code=3,angle=90,length=0.2)

```
Problem 1b --  same plot for the annual cycle phase

It has shifted approximately 7 days. 
```{r}

# pre-define arrays for regression parameters of seven data segments \nuline
ann1_phase <- 1:7
ann1_pherr <- 1:7
# ann2_phase <- 1:7
# ann2_pherr <- 1:7
# yr <- 1:7

k=0
# the data segments are defined by their start-points
for (j in c(1,109,217,321,429,537,637)){
  k=k+1
  timed <- time[j:(j+119)]
  time2d <- timed^2
  co2d <- co2[j:(j+119)]
  annsd <- anns[j:(j+119)]
  anncd <- annc[j:(j+119)]
  ann2sd <- ann2s[j:(j+119)]
  ann2cd <- ann2c[j:(j+119)]
  lmlmlm <- lm(co2d ~ timed + time2d + annsd + anncd + ann2sd + ann2cd)
  anns1 <- coef(summary(lmlmlm))["annsd","Estimate"]
  annc1 <- coef(summary(lmlmlm))["anncd","Estimate"]
  annamp <- sqrt(anns1^2 + annc1^2)
  ann1_phase[k] <- 180*atan2(anns1,annc1)/pi
  anns1 <- coef(summary(lmlmlm))["annsd","Std. Error"]
  annc1 <- coef(summary(lmlmlm))["anncd","Std. Error"]
  ann1_pherr[k] <- (180/pi)*sqrt(anns1^2+annc1^2)/annamp
}

dif <- ann1_phase[1]-ann1_phase[length(ann1_phase)]
dif
plot(yr,ann1_phase,col='black',xlab="Year",ylab="Annual Cycle Phase (degrees)",ylim=range(ann1_phase+ann1_pherr,ann1_phase-ann1_pherr))

```
Problem 1c -- plots for time evolution of semi-annual amplitude and phase -

The amplitude has increased since the 1960s. 
```{r}

# pre-define arrays for regression parameters of seven data segments
ann1 <- 1:7
ann1err <- 1:7
ann2 <- 1:7
ann2err <- 1:7
yr <- 1:7

anns <- sin(pi2*time)
annc <- cos(pi2*time)
ann2s <- sin(pi4*time)
ann2c <- cos(pi4*time)
# initialize segment counter
k=0
# the data segments are defined by their start-points
for (j in c(1,109,217,321,429,537,637)){
  k=k+1
  timed <- time[j:(j+119)]
  time2d <- timed^2
  co2d <- co2[j:(j+119)]
  annsd <- anns[j:(j+119)]
  anncd <- annc[j:(j+119)]
  ann2sd <- ann2s[j:(j+119)]
  ann2cd <- ann2c[j:(j+119)]
  lmlmlm <- lm(co2d ~ timed + time2d + annsd + anncd + ann2sd + ann2cd)
  summary(lmlmlm)
  anns2 <- coef(summary(lmlmlm))["ann2sd","Estimate"]
  annc2 <- coef(summary(lmlmlm))["ann2cd","Estimate"]
  ann2[k] <- sqrt(anns2^2+annc2^2) 
  yr[k] <- year[j+60]
  anns2 <- coef(summary(lmlmlm))["ann2sd","Std. Error"]
  annc2 <- coef(summary(lmlmlm))["ann2cd","Std. Error"]
  ann2err[k] <- sqrt(anns2^2+annc2^2)
}

plot(yr,ann2,col='black',xlab="Year",ylab="SemiAnnual Cycle Amplitude (ppm CO2)",ylim=range(ann2+ann2err,ann2-ann2err))
arrows(x0=yr,y0=ann2+ann2err,x1=yr,y1=ann2-ann2err,code=3,angle=90,length=0.2)

```
Problem 2a -- fit log(data) with a two-parameter model

```{r}

lm_logco2 <- lm(log(co2)~time)
pred_logco2 <- predict(lm_logco2)

plot(time,log(co2), type='l',xlab="")
lines(time, pred_logco2)
title("Mauna Loa Carbon Dioxide Measurements",xlab="year")

```
problem 2b -- log everything back and note predicted values in 1958 and 2021

The model under predicts the CO2 concentrations at the end points. In addition, it over predicts the middle of the data (1980s-2000s).
```{r}

exp_pred <- exp(pred_logco2)

print(paste("Predictions for March 1958: ",exp_pred[1],"ppm and August 2021: ",exp_pred[length(exp_pred)]))
print(paste("Actual Values for March 1958: ",co2[1],"ppm and August 2021: ",co2[length(exp_pred)]))

plot(time,co2,col='black',type="l",xlab="",ylab="")
lines(time, exp_pred)
title("Mauna Loa Carbon Dioxide Measurements",xlab="year",ylab="°Parts Per Million")
```

Problem 2c -- nonlinear least squares for 3-parameter exponential CO2 model -- test starting values of model parameters

The standard deviation of the residual variance is larger.
```{r}

s <- summary(lmlmlm)
sr <- s$sigma
sr


astart <- coef(summary(lm_logco2))["(Intercept)","Estimate"]
bstart <- co2[1]
cstart <- coef(summary(lm_logco2))["time","Estimate"]

time0 <- time-1788
nlmco2 <- nls(co2 ~ a+b*exp(c*time0),start=list(a=astart,b=bstart,c=cstart))
predexp <- predict(nlmco2)

s_nlm <- summary(nlmco2)
s_nlmr <- s_nlm$sigma
s_nlmr

print(paste(astart,bstart,cstart))

plot(time,co2,"l")
title(main="Mauna Loa Carbon Dioxide (ppm)")
lines(time,predexp)

aa <- coef(s_nlm)["a", "Estimate"]
aa
bb <- coef(s_nlm)["b", "Estimate"]
bb
cc <- coef(s_nlm)["c", "Estimate"]
cc

```
Problem 2d -- nonlinear least squares for 3-parameter exponential CO2 model 

```{r}
asd <- coef(s_nlm)["a", "Std. Error"]
t <- 1850-1958
x <- aa + bb*exp(cc*t)
x
dff <- 265 - aa
sgm <- dff/asd
sgm

```
Problem 2e

```{r}

t <- 1788-1958
x <- aa + bb*exp(cc*t)
x
t <- 0 
y <- aa + bb*exp(cc*t)
y

year_rate <- (y-x)/(1958-1788)
year_rate

dec_rate <- year_rate*10
dec_rate

```


problem 3a -- Singular value decomposition of the G-matrix, data representers referenced to the midpoint of the time series

```{r}

data_representer <- (time-1990.0)/25.0
data_representer2 <- data_representer^2
constant <- rep(1.0,length(data_representer))
gmatrix <- cbind(constant,data_representer,data_representer2,annc,anns,ann2c,ann2s)

gsvd <- svd(gmatrix,7,7)

singval <- gsvd$d
ksing <- seq(1, length(singval))

plot(ksing, singval, type = 'l')

condno <- max(singval)/min(singval)

print(paste("The condition number is ",condno))

```

Problem 3b -- print out the v-matrix singular vectors with largest and smallest singular values

The values are approximately in the same range.
```{r}

v <- gsvd$v

print("v-matrix singular vector with largest singular value")
v[1:7,1]

print("v-matrix singular vector with smallest singular value")
v[1:7,7]

```
problem 3c -- Singular value decomposition of the G-matrix, data representers referenced to the Birth of Christ

```{r}

data_representer <- time/2000.0
data_representer2 <- data_representer^2
constant <- rep(1.0,length(data_representer))
gmatrix <- cbind(constant,data_representer,data_representer2,annc,anns,ann2c,ann2s)

gsvd <- svd(gmatrix,7,7)

singval <- gsvd$d
ksing <- seq(1, length(singval))

plot(ksing, singval, type = 'l')

condno <- max(singval)/min(singval)

print(paste("The condition number is ",condno))

plot(time, data_representer, type='l')
lines(time, data_representer2)
lines(time, constant)


```
problem 4 -- bootstrap

```{r}

library(boot)
# define a function to obtain R-Squared from the data
model <- function(formula,data,indices){
  d <- data[indices,]
  fit <- lm(formula,data=d)
  return(coef(fit))
}

co2matrix <- data.frame(data=co2,g1=constant,g2=data_representer,g3=data_representer2,g4=annc,g5=anns,g6=ann2c,g7=ann2s)

# bootstrapping with 1000 replications 
results <- boot(data=co2matrix, statistic=model, 
   R=1000, formula=data~g2+g3+g4+g5+g6+g7)

results
plot(results)

rt <- results$t0
rt

```

