---
title: "Problem Set 2"
author: "Elizabeth Bailey"
date: "8/26/2021"
output:
  word_document: default
  pdf_document: default
  html_document: default
---
Question 1a 

a) For µ = 1 and σ = 2, plot the Gaussian PDF and CDF for −6 < X < 6.

```{r}

x <- seq(-6.0,6.0, by=0.01)
p <- seq(0.001,.999,by=0.001)

dd <- dnorm(x,mean=1, sd=2)
plot(x, dd, type='l')

pp <- pnorm(x,mean=1,sd=2)
plot(x, pp, type='l')

```
Question 1b

b) For µ = 1 and σ = 2, confirm the normalization condition R ∞−∞ p(X)dX = 1 with a numerical integral over an appropriate range of X with at least 1000 values of X and p(X).

```{r}

moo <- 1
sigma <- 2

x <- seq(-6.0,6.0, by=0.01)
y <- function(x) (1/(sqrt(2*pi)*sigma))*exp(-((x-moo)^2)/(2*sigma^2))

#Numerical integration
sum <- 0
for (x in seq(-100, 100, by=0.01)) {
  
  rum <- y(x)*.01
  sum <- sum + rum
}
sum

#Continuous Integration to check accuracy.
integrate(y, lower=-Inf, upper=Inf)
```
Quesiton 1c

c) For µ = 1 and σ = 2, confirm the mean of the PDF R ∞−∞ Xp(X)dX = µ with a numerical integral over an appropriate range of X with at least 1000 values of X and p(X).

```{r}
moo <- 1
sigma <- 2

x <- seq(-6.0,6.0, by=0.01)
meen <- function(x) {x*(1/(sqrt(2*pi)*sigma))*exp(-((x-moo)^2)/(2*sigma^2))}

#Numerical integration
sum <- 0
for (x in seq(-100, 100, by=0.01)) {
  
  rum <- meen(x)*.01
  sum <- sum + rum
}
sum

#Continuous Integration to check accuracy.
integrate(meen, lower=-Inf, upper=Inf)

```
Question 1d 

d) For µ = 1 and σ = 2, confirm the variance of the PDF R ∞−∞(X − µ)2p(X)dX = σ2 with a numerical integral over an appropriate range of X with at least 1000 values of X and p(X).

```{r}
moo <- 1
sigma <- 2

x <- seq(-6.0,6.0, by=0.01)
vary <- function(x) {(x-moo)^2*(1/(sqrt(2*pi)*sigma))*exp(-((x-moo)^2)/(2*sigma^2))}

#Numerical integration
sum <- 0
for (x in seq(-100, 100, by=0.01)) {
  
  rum <- vary(x)*.01
  sum <- sum + rum
}
sum

#Continuous Integration to check accuracy.
integrate(vary, lower=-Inf, upper=Inf)

```
Question 1e

e) For µ = 0 and σ = 1, the sample markdown file probset2 template.Rmd contains the following commands:
Explain why the sixth line of R code is useful: ddnn <- nn*dd/5.0.

It is a way of converting the function from the probability of the function to a function of the number of times an event occurs in the sample size (the y-axis converts from probabilities to frequencies).

```{r}
x <- seq(-5,5, by=.01) 
dd <- dnorm(x, sd=1)
plot(x,dd)
qq <- qnorm(p,sd=1)
plot(p,qq)
nn <- 10000
xx <- rnorm(nn,sd=1)
hist(xx,breaks=seq(-6,6,by=0.2))
ddnn <- nn*dd/5
lines(x,ddnn)
``` 
Question 1f

f) For the Gaussian with µ = 1 and σ = 2, what are the bounding values of the 10% and 90% percentiles of the distribution?

```{r}

moo <- 1
sig <- 2

tenperc <- qnorm(.10, mean=moo, sd=sig)
tenperc
nineperc <- qnorm(.90, mean=moo, sd=sig)
nineperc

```
Question 2a

a) For ν = 2, show that the χ2 PDF reduces to a normalized exponential function, that is, normalized so that Z ∞0 p(X)dX = 1

```{r}

nu <- 2
k <- nu/2
gma <- factorial(k)
pfx <- function(x) {(1/(2^(nu/2)*gma))*x^((nu/2)-1)*exp(-x/2)}

#Numerical integration
sum <- 0
for (x in seq(.0001, 1000, by=0.01)) {
  
  rum <- pfx(x)*.01
  sum <- sum + rum
}
sum

#Continuous Integration to check accuracy.
integrate(pfx, lower=0, upper=Inf)

```
Question 2b

b) For ν = 4, plot the PDF for X ∈ [0, 10], compute the peak values of the PDF using calculus and mark it on the graph.

```{r}

x <- seq(0, 10.0, by=0.01)
p <- seq(0.001,.999,by=0.001)

dchi <- dchisq(x, df=4)
plot(x,dchi,type='l')
lines(2, (1/(2*exp(1))), type='p', col='red')

```
Question 2c

c) For ν = 4, confirm the normalization condition R ∞ 0 p(X)dX = 1 with a numerical integral over an appropriate range of X with at least 1000 values of X and p(X). Compute the expected value hXi numerically as well, to confirm its theoretical value hXi = ν.

```{r}

nu <- 4
k <- nu/2
gma <- factorial(k-1)
pfx <- function(x) {(1/((2^(nu/2))*gma))*(x^((nu/2)-1))*(exp(-x/2))}

#Numerical integration
sum <- 0
for (x in seq(.0001, 1000, by=0.01)) {
  
  rum <- pfx(x)*.01
  sum <- sum + rum
}

sum

#Continuous Integration to check accuracy.
integrate(pfx, lower=0, upper=Inf)

men <- function(x) {x*(1/(2^(nu/2)*gma))*x^((nu/2)-1)*exp(-x/2)}

#Numerical integration
sum <- 0
for (x in seq(.0001, 1000, by=0.01)) {
  
  rum <- men(x)*.01
  sum <- sum + rum
}
sum

#Continuous Integration to check accuracy.
integrate(men, lower=0, upper=Inf)

```
Question 2d

d) For ν = 10, plot the PDF for X ∈ [0, 30], compute the peak values of the PDF using calculus and mark it on the graph.

```{r}

x <- seq(0, 30.0, by=0.01)
p <- seq(0.001,.999,by=0.001)

dchi <- dchisq(x, df=10)
plot(x,dchi, type='l')
lines(8, (16/(3*exp(4))), type='p', col='red')
```
Question 2e

e) For ν = 2,4,6,8, and 10, use the CDF function pchisq to compute the probability of X > 2ν for each value of ν. Do these relative deviations from the expected value hXi = ν become less probable or more probable as ν increases?

They become less probable as nu increases.
```{r}
c=matrix(0,5,2)

for (n in seq(2, 10, 2)) {

nu <- n
i <- n/2
c[i,1] <- n

q <- seq(0, 30.0, by=0.01)
pch <- pchisq(q, df=nu)
plot(q, pch, type='l')
par(new=TRUE)

p2nu <- pchisq(2*nu, df=nu)
prb <- 1 - p2nu
prb
c[i,2] <- prb

}
c
```
Question 3a

a) For µ = 0 and σ = 1, plot the log-normal PDF and CDF for 0 < X < 6. Limit the y axis if there is a singularity in the PDF as X → 0+.

```{r}

x <- seq(0,6,by=.01)
dl <- dlnorm(x, meanlog = 0, sdlog = 1)
plot(x, dl, type='l')

pl <- plnorm(x, meanlog = 0, sdlog = 1)
plot(x, pl, type='l')

```
Question 3b

b) For µ = 0 and σ = 1, compute the expected value of the mean hXi = R ∞ 0 Xp(X)dX with a numerical integral over an appropriate range of X with at least 1000 values of X and p(X).

```{r}
 
moo <- 0
sig <- 1

pfx <- function(x) {x/(sig*x*sqrt(2*pi))*exp(-((log(x)-moo)^2)/(2*(sig^2)))}

#Numerical integration
sum <- 0
for (x in seq(0.001, 1000, by=0.01)) {
  
  rum <- pfx(x)*.01
  sum <- sum + rum
}
sum

#Continuous Integration to check for accuracy
integrate(pfx, lower=0, upper=Inf)

```
Question 3c

c) For µ = 2 and σ = 1, plot the log-normal PDF and CDF for 0 < X < 25. Limit the y axis if there is a singularity in the PDF as X → 0+.

```{r}

x <- seq(0,25,by=.01)
dl <- dlnorm(x, meanlog = 2, sdlog = 1)
plot(x, dl, type='l')

pl <- plnorm(x, meanlog = 2, sdlog = 1)
plot(x, pl, type='l')

```
Question 3d

d) For µ = 2 and σ = 1, compute the expected value of the mean hXi = R ∞ 0 Xp(X)dX with a numerical integral over an appropriate range of X with at least 1000 values of X and p(X).

```{r}

moo <- 2
sig <- 1

pfx <- function(x) {x/(sig*x*sqrt(2*pi))*exp(-((log(x)-moo)^2)/(2*(sig^2)))}

#Numerical integration
sum <- 0
for (x in seq(0.001, 1000, by=0.01)) {
  
  rum <- pfx(x)*.01
  sum <- sum + rum
}
sum

#Continuous Integration to check for accuracy
integrate(pfx, lower=0, upper=Inf)

```
Question 4a

a) For N = 10, p = 0.5, plot the binomial PDF and CDF for 0 ≤ K ≤ 10. For N = 100, p = 0.5, plot the binomial PDF and CDF for 0 ≤ K ≤ 100.

```{r}

k <- seq(0, 10, by=1)
dbi <- dbinom(k, size = 10, prob = 0.5)
plot(k, dbi, type='l')

pbi <- pbinom(k, size = 10, prob = 0.5)
plot(k, pbi, type='l')

```
```{r}

k <- seq(0, 100, by=1)
dbi <- dbinom(k, size = 100, prob = 0.5)
plot(k, dbi, type='l')

pbi <- pbinom(k, size = 100, prob = 0.5)
plot(k, pbi, type='l')

```
Question 4b

b) For N = 10, p = 0.1, plot the binomial PDF and CDF for 0 ≤ K ≤ 10. For N = 100, p = 0.1, plot the binomial PDF and CDF for 0 ≤ K ≤ 100.

```{r}

k <- seq(0, 10, by=1)
dbi <- dbinom(k, size = 10, prob = 0.1)
plot(k, dbi, type='l')

pbi <- pbinom(k, size = 10, prob = 0.1)
plot(k, pbi, type='l')

```
```{r}

k <- seq(0, 100, by=1)
dbi <- dbinom(k, size = 100, prob = 0.1)
plot(k, dbi, type='l')

pbi <- pbinom(k, size = 100, prob = 0.1)
plot(k, pbi, type='l')

```

Question 4c

c) Assume that great earthquakes (Richter magnitude M ≥ 8.0) occur yearly in a plate boundary segment, say the Western Aleutians, with a fixed probability p = 0.01 events per year. Assume also that each year’s earthquakes are statistically independent of all other years’ earthquakes. In a single century, what is the probability of K ≥ 3 earthquakes on this plate boundary segment? In a single century, what is the probability of K ≥ 5 earthquakes on this plate-boundary segment?

```{r}

p <- 0.01 #probability of the event occurring
N <- 100 #Number of trials of the event
k <- seq(0, 100, by=1) #The number of instances of an event
dbi <- dbinom(4:100, size = N, prob = p)
prob3 <- sum(dbi)
prob3

dbi <- dbinom(6:100, size = N, prob = p)
prob5 <- sum(dbi)
prob5


```
Question 4d

d) Let p3 be the probability of K ≥ 3 earthquakes on the plate-boundary segment in part (c). The typical number of centuries that one might expect to occur in order to observe a single instance of K ≥ 3 great earthquakes is the reciprocal of this probability, that is, M3 ≈ (p3) −1 3 centuries. Use the function rbinom to realize 100 centuries of great-earthquake activity with these binomial statistics. How many centuries in your binomial realizations have K ≥ 3 events? Do these realizations comport with the probability you estimated in part (c)?

These realizations typically have 20 to 30 instances of K ≥ 3 great earthquakes occurring over 100 centuries, which is much more frequent than the probability estimated in part (C) (m3 ~ every 54 centuries).
```{r}

x <- seq(0, 100, by=1)
p3 <- prob3
m3 <- 1/p3
m3
rbi <- rbinom(x, size=N, prob=p3)
max(rbi)
plot(x, rbi)

count=0
for(i in 1:length(x)) {
if(rbi[i]>=3) {
count=count+1
}
}
count

```

