---
title: "Problem Set 3"
author: "Elizabeth Bailey"
date: "9/21/2021"
output:
  word_document: default
  pdf_document: default
  html_document: default
---

Question 1a

```{r}
N <- 9.0
n <- seq(1.0, N, by=1.0)
x <- n^2
print(x)

men <- sum(x)/N
print(men)

```
Question 1b -- median

```{r}

med <- x[5]
print(med)

```
Question 1c -- harmonic mean

```{r}
N <- 9.0
n <- seq(1.0, N, by=1.0)
x <- n^2

xinv <- x^{-1}
moohrecip <- sum(xinv)/N
mooh <- moohrecip^{-1}
mooh

```
Question 1d -- geometric mean

```{r}
N <- 9.0
n <- seq(1.0, N, by=1.0)
x <- n^2

xlog = log(x)
moogln <- sum(xlog)/N
moog <- exp(moogln)
moog
```
problem 2e
```{r}
M <- 5000
delx <- 0.01
# x <- (k-0.5)*delx
y <- function(x) ((x/4)*exp(-x/2))

#Numerical integration from 0 to 50
sum <- 0
for (k in seq(1, M, by=1)) {
  x <- (k-0.5)*delx
  rum <- log(x)*y(x)*delx
  sum <- sum + rum
}
sum
mg <- exp(sum)
mg

#Numerical integration from 0 to 100
M <- 10000
sum <- 0
for (k in seq(1, M, by=1)) {
  x <- (k-0.5)*delx
  rum <- log(x)*y(x)*delx
  sum <- sum + rum
}
sum
mg <- exp(sum)
mg

# #Continuous Integration to check accuracy.
# intgrd <- function(x) log(x)*y(x)
# integrate(intgrd, lower=0, upper=Inf)

x <- seq(.001, 100, by=.01)
intgrd <- log(x)*y(x)
plot(x,intgrd, type='l')
```
Problem 3a

```{r}

sunspots <- read.csv(file="sunspots_annual_1700_2008.csv")
head(sunspots)

yr <- sunspots$YEAR
sspots <- sunspots$SUNACTIVITY
plot(yr,sspots)

smax <- max(sspots)
smax
smin <- min(sspots)
smin

hist(sspots, breaks = 20, freq = TRUE)
```
Problem 3b
```{r}
sum <- sum(sspots)
N <- 2009-1700
men <- sum/N
men

mean(sspots)

dif <- (sspots-men)
difsq <- dif^2
syum <- sum(difsq)
sampstd <- sqrt(syum/N)
sampstd
```
Problem 3c

This PDF doesn't look good because the standard deviation seems to be off and the mean peaks at a different point. 

```{r}
x <- seq(smin, smax, by=1)

nn <- 10000
dd <- dnorm(x,mean=men, sd=sampstd)
ddnn <- nn*dd/3



plot(x, dd, type='l')

hist(sspots, breaks = 20, freq = TRUE)
lines(x,ddnn)

```
 Problem 3d
 
```{r}
 
lspots <- log(sspots+10.0)
lsmin <- min(lspots)
lsmin
lsmax <- max(lspots)
lsmax
 
hist(lspots, breaks=50)

sum <- sum(lspots)
sum
N <- 2009-1700
men <- sum/N
men

mean(lspots)

dif <- (lspots-men)
difsq <- dif^2
syum <- sum(difsq)
syum
sampstd <- sqrt(syum/N)
sampstd


```
Problem 3e
I think it is a better than part c, but the distribution in the histogram seems to be more even without much of a centralized peak.
```{r}

x <- seq(lsmin, lsmax, by=.01)
# x <- seq(smin, smax, by=1)

nn <- 10000
ddnn <- nn*dd/3


dd <- dnorm(x,mean=men, sd=sampstd)
plot(x, dd, type='l')

hist(lspots, breaks = 50, freq = TRUE)

```
 Problem 3f
 
```{r}
 
sortspots <- sort(sspots)
N <- length(sortspots)

smax <- max(sspots)
smax
smin <- min(sspots)
smin

m <- seq(smin, smax, by=1)
kl <- length(m)
kl

empcdf = matrix(0,kl,1)


sum <- 0
for (k in seq(smin, smax, by=1)) {
  ss <- sortspots>=k & sortspots<(k+1)
  n <- length(ss[ss == TRUE])
  ratio <- n/N
  sum <- sum + ratio
  empcdf[k+1] <- sum
}
head(empcdf)
plot(m, empcdf, type='l')
title("empirical CDF of sunspot data 1700AD-2009AD")

```
  Problem 3g
  
  
```{r}

plot(log(sspots))

```
```{r}
moo <- 3.5
sgm <- 0.9
S <- seq(.001,1000, by=1)

ps <- function(S) ((1/(sgm*sqrt(2*pi)*(10+S)))*exp(-(log(10+S)-moo)^(2)/(2*sgm^(2))))

integrate(ps, lower = 0, upper = Inf)

#Numerical integration
sum <- 0
cdf = matrix(0,length(S),1)
for (i in seq(.001,1000, by=1)) {
  c <- ps(i)
  sum <- sum + c
  cdf[i+1] <- sum
}
sum

plot(cdf, type='l', ylim = c(0, 1))
lines(m, empcdf, type='l', col='red')

```

```{r}
moo <- 3.75
sgm <- .85
S <- seq(.001,1000, by=1)

ps <- function(S) ((1/(sgm*sqrt(2*pi)*(10+S)))*exp(-(log(10+S)-moo)^(2)/(2*sgm^(2))))

integrate(ps, lower = 0, upper = Inf)

#Numerical integration
sum <- 0
cdf = matrix(0,length(S),1)
for (i in seq(.001,1000, by=1)) {
  c <- ps(i)
  sum <- sum + c
  cdf[i+1] <- sum
}
sum

plot(cdf, type='l', ylim = c(0, 1))
lines(m, empcdf, type='l', col='red')

```

```{r}
moo <- 4
sgm <- 0.8
S <- seq(.001,1000, by=1)

ps <- function(S) ((1/(sgm*sqrt(2*pi)*(10+S)))*exp(-(log(10+S)-moo)^(2)/(2*sgm^(2))))

integrate(ps, lower = 0, upper = Inf)

#Numerical integration
sum <- 0
cdf = matrix(0,length(S),1)
for (i in seq(.001,1000, by=1)) {
  c <- ps(i)
  sum <- sum + c
  cdf[i+1] <- sum
}
sum

plot(cdf, type='l', ylim = c(0, 1))
lines(m, empcdf, type='l', col='red')

```


  Problem 4 -- negative binomial
  
  Problem 4a
  
```{r}

k <- 4
s <- 0.5
N <- seq(4,20,by=1)

nbin <- choose(N-1, k-1)*s^(k)*(1-s)^(N-k)

plot(N, nbin, type='l')
  
```
  Problem 4b
  
```{r}

nbpdf <- dnbinom(N,k,s)
plot(N, nbpdf, type='l')
title("Probability of four heads after a number of coin flips")
  
nbcdf <- pnbinom(N,k,s)
plot(N, nbcdf, type='l')
title("Cumlative probability of four heads after a number of coin flips")

```
  Problem 4c

```{r}
k <- 4
s <- 0.1
N <- seq(k,100,by=1)

nbpdf <- dnbinom(N,k,s)
plot(N, nbpdf, type='l')
title("Probability of four heads after a number of coin flips")
  
nbcdf <- pnbinom(N,k,s)
plot(N, nbcdf, type='l')
title("Cumlative probability of four heads after a number of coin flips")
```
