---
title: "Problem Set 6"
author: "Elizabeth Bailey"
date: "9/30/2021"
output:
  word_document: default
  pdf_document: default
  html_document: default
---

data files converted to CSV files with these commands

echo 'YEAR, MONTH, TIME, CO2 ' >! co2_maunaloa.csv
awk 'NR>53 {print $1 ", " $2 ", " $3 ", " $4}' co2_mm_mlo.txt >> co2_maunaloa.csv


Question 1a -- linear trend fit to Mauna Loa CO2

The model underestimates the beginning and end of the data, but it captures the middle part of the data well.

```{r}

mlo_co2 <- read.csv("co2_maunaloa.csv")
temps <- mlo_co2$DTEMPC
years <- mlo_co2$YEAR
months <- mlo_co2$MONTH
time <- mlo_co2$TIME
co2 <- mlo_co2$CO2

# title("Mauna Loa Carbon Dioxide Measurements",xlab="year",ylab="°Parts Per Million")

lmco2 <- lm(co2 ~ time)
summary(lmco2)
yrly <- lmco2$coefficients[2]
yrly
dcadl <- yrly*10
dcadl
plot(lmco2)

```
Problem 1b -- start and finish CO2 predictions

both end points are underestimated by the model.
```{r}

predco2 <- predict(lmco2)
start <- predco2[1]
end <- predco2[length(predco2)]
start0 <- co2[1]
end0 <- co2[length(co2)]

print(paste("Starting value of linear model is ",start," ppm in March 1958, and finish is ",end," ppm in August 2021"))

print(paste("Starting value of actual data is ",start0," ppm in March 1958, and finish is ",end0," ppm in August 2021"))
```
Problem 1c --  
```{r}

residual1 <- (lmco2$residuals)
s <- summary(lmco2)
s$fstatistic[1]

print(paste("The standard deviation of the data misfit to the linear model is ", sd(residual1)," ppm"))

```
problem 1d -- plot co2 data against the linear model

The rate of change is overestimated at the beginning of the curve and underestimated at the end of the curve.
```{r}

lmco2 <- lm(co2 ~ time)
predco2 <- predict(lmco2)
plot(time,co2,"l")
lines(time,predco2)
title("Mauna Loa CO2 versus linear-ramp model")

```

Problem 1e -- fit the data with a quadratic

The start and end values are pretty good representations of the data - better than the linear model.
```{r}

time2 <- time^2
lm2co2 <- lm(co2~time+time2)
# summary(lm2co2)
pred2co2 <- predict(lm2co2)

start <- pred2co2[1]
finish <- pred2co2[length(pred2co2)]
start0 <- co2[1]
finish0 <- co2[length(co2)]

print(paste("Starting value of quadratic model is ",start," ppm in March 1958, and finish is ",finish," ppm in August 2021"))

print(paste("Starting value of actual data is ",start0," ppm in March 1958, and finish is ",finish0," ppm in August 2021"))



```
problem 1f -- yearly increases of the quadratic model at start and finish of CO2 series

```{r}

start_ramp <- pred2co2[13]-pred2co2[1]
finish_ramp <- pred2co2[762]-pred2co2[750]

print(paste("Starting CO2 increase per year of quadratic model is ",start_ramp," ppm in 1958-1959, and ending CO2 increase per year is ",finish_ramp," ppm in 2020-2021"))

```
problem 1g -- standard deviation of the misfit to data

```{r}

residual2 <- (lm2co2$residuals)
s <- summary(lm2co2)
s$fstatistic[1]

print(paste("The standard deviation of the data misfit to the quadratic model is ", sd(residual2)," ppm"))

```

problem 1h -- plot the data against quadratic model

The model does not capture the seasonal cycle of the data.
```{r}

plot(time,co2,"l")
lines(time,pred2co2)
title("Mauna Loa CO2 versus quadratic model")

```
problem 1i -- cubic model fit

There is no significant change to the fit of the model from the cubic to quadtradic models. There is not a big change in the standard deviation of the data misfit.

```{r}

time2 <- time^2
time3 <- time^3
lm3co2 <- lm(co2~time+time2+time3)
# summary(lm3co2)

pred3co2 <- predict(lm3co2)

start <- pred3co2[1]
finish <- pred3co2[length(pred3co2)]
start0 <- co2[1]
finish0 <- co2[length(co2)]

residual3 <- (lm3co2$residuals)
s <- summary(lm3co2)
s$fstatistic

print(paste("Starting value of cubic model is ",start," ppm in March 1958, and finish is ",finish," ppm in August 2021"))

print(paste("Starting value of actual data is ",start0," ppm in March 1958, and finish is ",finish0," ppm in August 2021"))

print(paste("The standard deviation of the data misfit to the cubic model is ", sd(residual3)," ppm"))

plot(time,co2,"l")
lines(time,pred3co2)
title("Mauna Loa CO2 versus cubic model")

```
Problem 2a -- add the annual cycle to the co2 datafit

```{r}
pi2 <- 2*pi
anns <- sin(pi2*time)
annc <- cos(pi2*time)
lm2aco2 <- lm(co2 ~ time + time2 +anns + annc)

coefs <- coef(summary(lm2aco2))["anns","Estimate"]
coefc <- coef(summary(lm2aco2))["annc","Estimate"]
coefann <- sqrt(coefs^2 + coefc^2)
dcoefs <- coef(summary(lm2aco2))["anns","Std. Error"]
dcoefc <- coef(summary(lm2aco2))["annc","Std. Error"]
dcoefann <- sqrt(dcoefs^2 + dcoefc^2)

print(paste("The amplitude of the full annual cycle is ",coefann," ppm with standard deviation ",dcoefann," ppm"))

```
Problem 2b -- Annual cycle amplitude
Problem 2b -- Annual cycle phase

```{r}

phi <- atan2(coefs,coefc)
peakyr <- phi/(2*pi)
peakmon <- peakyr*12

print(paste("peak phase at the year fraction ",peakyr,"  month index is ",peakmon))

```
Problem 2c -- plot residuals of datafit without and with the annual-cycle fit

```{r}

resid <- co2 - pred2co2
residual21 <- co2 - predict(lm2aco2)

print(paste("The st dev of the data misfit to the quadratic model is ", sd(resid)," ppm"))
print(paste("The st dev of the data misfit to the quadratic model plus annual cycle is ", sd(residual21)," ppm"))

plot(resid,residual21)
title("Mauna Loa CO2 series after least squares fits")

plot(resid, type='l')
lines(residual21, type='l', col='yellow')
```
problem 2d -- fitting a bi-annual term

```{r}

pi4 <- 4*pi
ann2s <- sin(pi4*time)
ann2c <- cos(pi4*time)
lm2a2co2 <- lm(co2 ~ time + time2 + anns + annc + ann2s + ann2c)

coef2s <- coef(summary(lm2a2co2))["ann2s","Estimate"]
coef2c <- coef(summary(lm2a2co2))["ann2c","Estimate"]
coef2ann <- sqrt(coef2s^2 + coef2c^2)
dcoef2s <- coef(summary(lm2a2co2))["ann2s","Std. Error"]
dcoef2c <- coef(summary(lm2a2co2))["ann2c","Std. Error"]
dcoef2ann <- sqrt(dcoef2s^2 + dcoef2c^2)

phi <- atan2(coef2s,coef2c)
peakyr <- phi/(2*pi)
peakmon <- peakyr*12

print(paste("peak phase at the year fraction ",peakyr,"  month index is ",peakmon))

print(paste("The amplitude of the semi-annual cycle is ",coef2ann," ppm with standard deviation ",dcoef2ann," ppm"))

plot(time, predict(lm2a2co2), type = 'l')
lines(time, predict(lm2aco2), type='l', col='yellow')
title("Mauna Loa CO2 series annual cycles")

```
Problem 2e -- plot residuals of datafit with annual cycle and with the adjusted annual-cycle fit

It add the seasonal cycle to the model. 

```{r}

residual22 <- co2 - predict(lm2a2co2)
print(paste("The st dev of the data misfit to the quadratic model plus adjusted annual cycle is ", sd(residual22)," ppm"))

plot(residual22, type='l')
title("Mauna Loa CO2 series after least squares fits")
```
